Two subtle problems with over-representation analysis

### 2024-06-07 manuscript/manuscript.Rmd
ms updates

>>>>

### Quantifying the effect of these problems on precision and recall using simulated expression data

RNA-seq counts for SRA run accession ERR2539161 were obtained from DEE2 [@Mo2018-zb].
This dataset was selected due to its high number of assigned reads (367M).
A library of 200 random gene sets were generated, each containing 30 genes.
To generate differential expression profiles, 5% of gene sets were selected to be differentially
expressed, with equal numbers of up- and down-regulated pathways.
Pseudosamples were generated by first downsampling the counts using the `thincounts` function of
the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the addition of some extra variation drawn from a
normal distribution with mean of 1 and standard deviation varying between 0.3 and 0.45, this
value selected randomly for each gene.
Five "control" and five "case" pseudosamples were generated.
For the genes selected to be differentially expressed, expression values were multiplied by a
fold change in the case samples of 2.0 for upregulated genes and 0.5 for downregulated genes.
Genes that were selected to be both up and downregulated were left unchanged.

With these expression profiles, DESeq2 was used for differential analysis, and the results were
subjected to `enricher` which suffers from both problems, `fora`, an ORA function of the `fgsea`
bioconductor package v1.30.0 which does not suffer these problems, and `fgsea` a fast
implementation of the preranked GSEA algorithm [@Korotkevich2016-gd].
For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
selected for the foreground gene set.
If fewer than 250 genes are detected as significant, then 250 genes with the smallest p-values
were used for ORA, similar to a previous report [@Tarca2013-xh].
For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.

After enrichment analysis with these approaches, the significant sets up and down regulated sets
were selected and compared to the ground truth in order to calculate precision, recall and F1
score.

At each value of added variance, 500 replications were conducted, each with as set seed for
reproducibility.
